Demographic Species Distribution Models

### 2019-10-02 manuscript/ms.Rmd
add citations into old text

By contrast, matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a Leslie matrix representing key demographic processes13.
Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing Leslie matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the species’ distribution14.

@@

For each location, these rates were used to generate a 2x2 Leslie matrix and iterated to estimate the intrinsic growth rate $\lambda$.
As in a dynamic range model, species abundance was assumed to be proportional to $\lambda$ (this simplistic assumption will be improved upon in the proposed project) and probability of presence at each location derived via the complementary log-log link, with a spatially-constant detection rate parameter.
The model specification was completed with a Bernoulli likelihood and non-informative priors, and fitted to 500 presence-absence records from the Breeding Bird Survey of North America in 2015 [@todo].

@@

Basic DSDMs, like the model in Example 1, do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
Where such information is available (e.g. from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.
The DSDMs I propose are based on Leslie matrices, which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories23.
Like other Leslie matrix approaches, the basic DSDMs I have already developed can also be extended to explicitly model three key ecological processes that are poorly captured by existing distribution models: density dependence, dispersal, and biotic interactions.

@@

Density dependence can be represented in DSDMs by modelling demographic rates as a function of the current population state23 (the abundances of each life stage) as well as covariates.
For example, the current density of individuals could determine fecundity for the next iteration of the dynamics. The matrix iterations will then converge to a stable state; the population carrying capacity or the expected abundance.

@@

Dispersal can be incorporated in DSDMs by representing an entire metapopulation in a single sparse, block-structured Leslie matrix, with dispersal represented by transition probabilities between states in different patches24.

@@

Biotic interactionsbetween species can be incorporated in a DSDM by representing the interacting dynamics of multiple species in a single Leslie matrix, with inter-species density dependence25.
By modelling demographic rates of one species as dependent on the abundance of another species, DSDMs can represent the effects of competition, facilitation, predation and parasitism.

@@

Like many ecologically-realistic models, DSDM parameters are in general not uniquely identified, so estimates may be correlated or even multimodal.
This non-identifiability is ecologically informative as it arises from ignorance of the species’ ecology and highlights areas where more research is needed.
This ignorance can be captured in prediction uncertainty, more realistically representing how well we can predict distributions15.

@@

However, poor identification complicates statistical inference as maximum likelihood methods will be inaccurate and Markov chain Monte Carlo samplers may not converge16.
This proposal uses two new tools to resolve this issue and fit DSDMs to large datasets: Hamiltonian Monte Carlo & high performance automatic differentiation.

@@

Hamiltonian Monte Carlo (HMC) is a Markov chain Monte Carlo method that can sample effectively from highly correlated and multi-modal parameter distributions16, but requires gradient calculations.
Automatic differentiation (AD, also known as back-propagation) is a computational technique to derive these gradients. Whilst HMC & AD modelling software is available20, it is not optimised for the large-scale linear algebra required for DSDMs.
High-performance AD software enables complex machine learning models (e.g. deep neural networks) to be fit to very large datasets; facilitating recent dramatic advances in artificial intelligence.
Google’s recently released open source, high performance AD software TensorFlow is optimised for massive linear algebra21.
I have combined HMC with TensorFlow in an R package greta19 for flexible specification of and sampling from statistical models in R.
Using greta, ecologists can easily build and fit DSDMs on desktop computers or high-performance compute clusters.

>>>>

By contrast, matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a transition matrix representing key demographic processes [@merowAdvancingPopulationEcology2014].
Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing transition matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the species’ distribution [@merowUsingIntegralProjection2014].

@@

For each location, these rates were used to generate a 2x2 transition matrix (a Lefkovitch matrix) and iterated to estimate the intrinsic growth rate $\lambda$.
As in a dynamic range model, species abundance was assumed to be proportional to $\lambda$ (this simplistic assumption will be improved upon in the proposed project) and probability of presence at each location derived via the complementary log-log link, with a spatially-constant detection rate parameter.
The model specification was completed with a Bernoulli likelihood and non-informative priors, and fitted to 500 presence-absence records from the North American Breeding Bird Survey in 2015 [@sauerNorthAmericanBreeding2011].

@@

Basic DSDMs, like the model in Example 1, do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
Where such information is available (e.g. from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.
The DSDMs I propose are based on transition matrices (including the matrices used in both MPMs and IPMs), which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories [@ellnerIntegralProjectionModels].
Like other transition matrix approaches, the basic DSDMs I have already developed can also be extended to explicitly model three key ecological processes that are poorly captured by existing distribution models: density dependence, dispersal, and biotic interactions.

@@

Density dependence can be represented in DSDMs by modelling demographic rates as a function of the current population state [@ellnerIntegralProjectionModels] (the abundances of each life stage) as well as covariates.
For example, the current density of individuals could determine fecundity for the next iteration of the dynamics. The matrix iterations will then converge to a stable state; the population carrying capacity or the expected abundance.

@@

Dispersal can be incorporated in DSDMs by representing an entire metapopulation in a single sparse, block-structured transition matrix, with dispersal represented by transition probabilities between states in different patches [@neubertDEMOGRAPHYDISPERSALCALCULATION2000].

@@

Biotic interactionsbetween species can be incorporated in a DSDM by representing the interacting dynamics of multiple species in a single transition matrix, with inter-species density dependence [@fujiwaraCoexistenceCompetingStagestructured2011].
By modelling demographic rates of one species as dependent on the abundance of another species, DSDMs can represent the effects of competition, facilitation, predation and parasitism.

@@

Like many ecologically-realistic models, DSDM parameters are in general not uniquely identified, so estimates may be correlated or even multimodal.
This non-identifiability is ecologically informative as it arises from ignorance of the species’ ecology and highlights areas where more research is needed.
This ignorance can be captured in prediction uncertainty, more realistically representing how well we can predict distributions [@bealeIncorporatingUncertaintyPredictive2012].

@@

However, poor identification complicates statistical inference as maximum likelihood methods will be inaccurate and Markov chain Monte Carlo samplers may not converge [@monnahanFasterEstimationBayesian2017].
This proposal uses two new tools to resolve this issue and fit DSDMs to large datasets: Hamiltonian Monte Carlo & high performance automatic differentiation.

@@

Hamiltonian Monte Carlo (HMC) is a Markov chain Monte Carlo method that can sample effectively from highly correlated and multi-modal parameter distributions [@monnahanFasterEstimationBayesian2017], but requires gradient calculations.
Automatic differentiation (AD, also known as back-propagation) is a computational technique to derive these gradients. Whilst HMC & AD modelling software is available [@carpenterStanProbabilisticProgramming2017], it is not optimised for the large-scale linear algebra required for DSDMs.
High-performance AD software enables complex machine learning models (e.g. deep neural networks) to be fit to very large datasets; facilitating recent dramatic advances in artificial intelligence.
Google’s recently released open source, high performance AD software TensorFlow is optimised for massive linear algebra [@abadiTensorFlowLargeScaleMachine2016].
I have combined HMC with TensorFlow in an R package greta [@goldingGretaSimpleScalable2019] for flexible specification of and sampling from statistical models in R.
Using greta, ecologists can easily build and fit DSDMs on desktop computers or high-performance compute clusters.
